Aspects of Dealing with Imperfect Data in Temporal Databases

### 2012-03-16 book/sources/fuzzyTemporalDatabases.tex
update

%The time when a package would arrive at its destination may be estimated and can eventually be researched, but it is typically not always precisely known. For such companies and in many other situations, information systems able to handle imperfection corresponding to certain temporal aspects of the objects modelled by the system are necessary. The transition between granularities~\cite{Lin97} is also sometimes considered as a source of imprecision. Therefore, some proposals consider granularity as the base of the temporal model~\cite{Cru97}.
%Some software applications deal with uncertainty related to time. e.g. a logistics company which transports packages from one destination to another. The time when the package would be at the end point may be estimated but not precisely known. These applications need to manage the uncertainty with respect to the temporal attributes of the objects stored in the database.

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
A possibilistic variable $X$ over a universe $U$ is defined as a variable taking exactly one value in $U$, but for which this value is (partially) unknown. Its possibility distribution $\pi_X$ on $U$ models the available knowledge about the value that $X$ takes: for each $u\in U$, $\pi_X(u)$ represents the possibility that $X$ takes the value $u$. In the presented work, this possibility is interpreted as a measure of how plausible it is that $X$ takes the value $u$, given (partial) knowledge about the value $X$ takes.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

\subsection{Bipolarity in Temporal Databases}
\label{subsubsec:bipolarity}
Humans express their preferences using both positive and negative statements, where positive statements express what is desired or acceptable and negative statements express what is undesired or unacceptable \cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. This realization is interesting with regard to database querying, because sometimes a user does not exactly know his or her preferences or can't express them in only positive statements, but prefers to use negative statements to express what he or she dislikes or doesn't need. This introduces the need for \emph{bipolar querying}, a technique to model both positive and negative user preferences in a database query. Sometimes positive and negative preferences are clearly symmetric, making it possible to derive one from the other. For example, a person may define the concept of `tall' as `1.80 meters or higher'. The negative would then be the opposite: not tall would be `anyhing less than 1.80 meters'. However, in some cases, positive preferences can not be directly obtained from negative preferences or vice versa. E.g. when a person prefers to buy a black motorbike, this does not necessarily mean the person would totally reject a very dark blue motorbike. This phenomenon is called \emph{heterogeneous bipolarity}~\cite{Dubois2006},~\cite{Dubois2008}.

>>>>

%The time when a package would arrive at its destination may be estimated and can eventually be researched, but it is typically not always precisely known. For such companies and in many other situations, information systems able to handle imperfection corresponding to certain temporal aspects of the objects modelled by the system are necessary. The transition between granularities~\cite{Lin97} is also sometimes considered as a source of imprecision. Therefore, some proposals consider granularity as the base of the temporal model~\cite{Cru97}.
%Some software applications deal with uncertainty related to time. e.g., a logistics company which transports packages from one destination to another. The time when the package would be at the end point may be estimated but not precisely known. These applications need to manage the uncertainty with respect to the temporal attributes of the objects stored in the database.

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Possibilistic variable}~\cite{Pon11}\\
A possibilistic variable $X$ over a universe $U$ is defined as a variable taking exactly one value in $U$, but for which this value is (partially) unknown. Its possibility distribution $\pi_X$ on $U$ models the available knowledge about the value that $X$ takes: for each $u\in U$, $\pi_X(u)$ represents the possibility that $X$ takes the value $u$. In the presented work, this possibility is interpreted as a measure of how plausible it is that $X$ takes the value $u$, given (partial) knowledge about the value $X$ takes.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

\subsection{Bipolarity in Temporal Databases}
\label{subsubsec:bipolarity}
Humans express their preferences using both positive and negative statements, where positive statements express what is desired or acceptable and negative statements express what is undesired or unacceptable \cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. This realization is interesting with regard to database querying, because sometimes a user does not exactly know his or her preferences or can't express them in only positive statements, but prefers to use negative statements to express what he or she dislikes or doesn't need. This introduces the need for \emph{bipolar querying}, a technique to model both positive and negative user preferences in a database query. Sometimes positive and negative preferences are clearly symmetric, making it possible to derive one from the other. For example, a person may define the concept of `tall' as `1.80 meters or higher'. The negative would then be the opposite: not tall would be `anything less than 1.80 meters'. However, in some cases, positive preferences can not be directly obtained from negative preferences or vice versa. e.g., when a person prefers to buy a black motorbike, this does not necessarily mean the person would totally reject a very dark blue motorbike. This phenomenon is called \emph{heterogeneous bipolarity}~\cite{Dubois2006},~\cite{Dubois2008}.

### 2012-03-16 book/sources/timeDomain.tex
To my opinion, the content is now perfect.
update

Data models used by information systems (and in specific, temporal database systems) may represent these underlying time domains using \emph{chronons}~\cite{Dyreson1994}, which can informally be discribed as the smallest distinguishable time units that can be used in the system. However, to explain what chronons are, an explanation of some other temporal concepts is necessary.

@@

Orthogonal to the classification of underlying time models as linear or cyclic, they can be classified as \emph{discrete}, \emph{dense} or \emph{continuous} models\cite{Dyreson1994}. In a discrete model~\cite{Clifford:1985:AHR:971699.318922}, the notion exists that every instant has a unique successor and the set of instants is seen as a discrete one. Here, intuitively, the set of instants can be seen as isomorphic to the set of natural numbers $\Nat$. In a dense model, the notion exists that between any two instants always lies another. Here, intuitively, the set of instants can be seen as isomporphic to the set of rational numbers $\Q$ (when the set of instants is a discrete one) or the set of real numbers $\R$ (when the set of instants is a continuous one). In a continuous model, the notion also exists that between any two instants always lies another one, but the set of instants is always seen as continuous and there are no ``gaps'' between successive instants.

Some other necessary concepts are ~\cite{Dyreson1994}:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Time Interval}\\
A \emph{\textbf{time interval}} is the time between two instants.
\end{definition}

\begin{definition}\textbf{Duration}\\
A \emph{\textbf{duration}} is an amount of time with known length, but no specific starting or ending instants.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Temporal Element}\\
A \emph{\textbf{temporal element}} is a finite union of time intervals.
\end{definition}

\begin{definition}\textbf{Event}\\
An \emph{\textbf{event}} is an instantaneous fact, i.e. something occurring at an instant.
\end{definition}

\begin{definition}\textbf{`Temporal' as Modifier}\\
The modifier \emph{\textbf{`temporal'}} is used to indicate that the modified concept concerns some aspect of time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Chronon}\\
In a data model, a \emph{\textbf{chronon}} is a non-decomposable time interval of some fixed, minimal duration.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

Another classification of time models concerns the use of points or intervals to model time. The equivalence between interval-based and point-based time models is demonstrated in \cite{Böhlen_point-versus}.

@@

%\textcolor{red}{MOVE TO TEMPORAL DATABASE PART}
%On the other hand, time may be~\cite{Jensen94thetsql2} \textbf{relative} or \textbf{absolute} (in other words, \textbf{anchored} or \textbf{unanchored}). Sometimes, what one considers \emph{absolute time} is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g. -3 days = three days ago).
%\textcolor{red}{END MOVE}

@@

\begin{definition}\textbf{Granule}\\
A \textbf{\emph{granule}} is the basic time unit in a granularity.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Finner Than}\\
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{finner than}} granularity $H$ if:
\begin{equation}
\label{eq:finner-than}
| f \left( i_G \right) |\ <\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}

\begin{definition}
\label{def:coarser-than}\textbf{Coarser Than}
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{coarser than}} granularity $H$ if:
\begin{equation}
\label{eq:coarser-than}
| f \left( i_G \right)|\ >\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

Thus, the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Orthogonal to this classification, mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}~\cite{Lin97},~\cite{DyresonSnodgrass1994}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function value is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. E.g., the mapping value of the mapping between hours and minutes is calculated using a multiplication by $60$. 
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function value can not be calculated by means of multiplications and/or divisions. E.g., the mapping value of the mapping between months and days is dependent on the exact month or day. 
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities involved in the mapping have the same granules but a different anchor. E.g., the mapping between the days (Gregorian calendar days) and the academic days is a congruent mapping.
\end{itemize}

@@

%This subsection will concern \emph{calendars}. A calendar provides some sort of human interpretation of time. As such, calendars provide meaning or interpretation to temporal values where this meaning or interpretation is relevant or useful to the user. More precisely, calendars determine the mapping between human-meaningful time values and an underlying time model~\cite{Dyreson1994}. More formally, a calendar is an organization of different granularities (e.g. granularities day, month and year) in a hierarchy. In order to formally define what a calendar is, some other concepts should first be introduced~\cite{Kraus1997}:

@@

\subsection{Temporal Relationships}
%an explanation on allen's temporal relations.
%definition of fuzzy temporal interval
In this section, a brief introduction can be found, concerning \emph{temporal relationships}, sometimes also called `temporal relations'\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. Temporal relationships can be seen as relationships between temporal elements belonging to the same time domain. These relationships express how the temporal elements are related to one another, with respect to temporal precedence and overlap.

>>>>

Data models used by information systems (and in specific, temporal database systems) may represent these underlying time domains using \emph{chronons}~\cite{Dyreson1994}, which can informally be described as the smallest distinguishable time units that can be used in the system. However, to explain what chronons are, an explanation of some other temporal concepts is necessary.

@@

Orthogonal to the classification of underlying time models as linear or cyclic, they can be classified as \emph{discrete}, \emph{dense} or \emph{continuous} models\cite{Dyreson1994}. In a discrete model~\cite{Clifford:1985:AHR:971699.318922}, the notion exists that every instant has a unique successor and the set of instants is seen as a discrete one. Here, intuitively, the set of instants can be seen as isomorphic to the set of natural numbers $\Nat$. In a dense model, the notion exists that between any two instants always lies another. Here, intuitively, the set of instants can be seen as isomorphic to the set of rational numbers $\Q$ (when the set of instants is a discrete one) or the set of real numbers $\R$ (when the set of instants is a continuous one). In a continuous model, the notion also exists that between any two instants always lies another one, but the set of instants is always seen as continuous and there are no ``gaps'' between successive instants.

Some other necessary concepts are:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Time Interval}~\cite{Dyreson1994}\\
A \emph{\textbf{time interval}} is the time between two instants.
\end{definition}

\begin{definition}\textbf{Duration}~\cite{Dyreson1994}\\
A \emph{\textbf{duration}} is an amount of time with known length, but no specific starting or ending instants.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Temporal Element}~\cite{Dyreson1994}\\
A \emph{\textbf{temporal element}} is a finite union of time intervals.
\end{definition}

\begin{definition}\textbf{Event}~\cite{Dyreson1994}\\
An \emph{\textbf{event}} is an instantaneous fact, i.e. something occurring at an instant.
\end{definition}

\begin{definition}\textbf{`Temporal' as Modifier}~\cite{Dyreson1994}\\
The modifier \emph{\textbf{`temporal'}} is used to indicate that the modified concept concerns some aspect of time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Chronon}~\cite{Dyreson1994}\\
In a data model, a \emph{\textbf{chronon}} is a non-decomposable time interval of some fixed, minimal duration.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

Another classification of time models concerns the use of points or intervals to model time. The equivalence between interval-based and point-based time models is demonstrated in \cite{Bohlen1998}.

@@

%\textcolor{red}{MOVE TO TEMPORAL DATABASE PART}
%On the other hand, time may be~\cite{Jensen94thetsql2} \textbf{relative} or \textbf{absolute} (in other words, \textbf{anchored} or \textbf{unanchored}). Sometimes, what one considers \emph{absolute time} is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g., -3 days = three days ago).
%\textcolor{red}{END MOVE}

@@

\begin{definition}\textbf{Granule}~\cite{wang93}\\
A \textbf{\emph{granule}} is the basic time unit in a granularity.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Finner Than}~\cite{Lin97}\\
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{finner than}} granularity $H$ if:
\begin{equation}
\label{eq:finner-than}
| f \left( i_G \right) |\ <\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}

\begin{definition}
\label{def:coarser-than}\textbf{Coarser Than}~\cite{Lin97}\\
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{coarser than}} granularity $H$ if:
\begin{equation}
\label{eq:coarser-than}
| f \left( i_G \right)|\ >\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

Thus, the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Orthogonal to this classification, mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}~\cite{Lin97},~\cite{DyresonSnodgrass1994}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function value is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. e.g., the mapping value of the mapping between hours and minutes is calculated using a multiplication by $60$. 
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function value can not be calculated by means of multiplications and/or divisions. e.g., the mapping value of the mapping between months and days is dependent on the exact month or day. 
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities involved in the mapping have the same granules but a different anchor. e.g., the mapping between the days (Gregorian calendar days) and the academic days is a congruent mapping.
\end{itemize}

@@

%This subsection will concern \emph{calendars}. A calendar provides some sort of human interpretation of time. As such, calendars provide meaning or interpretation to temporal values where this meaning or interpretation is relevant or useful to the user. More precisely, calendars determine the mapping between human-meaningful time values and an underlying time model~\cite{Dyreson1994}. More formally, a calendar is an organization of different granularities (e.g., granularities day, month and year) in a hierarchy. In order to formally define what a calendar is, some other concepts should first be introduced~\cite{Kraus1997}:

@@

\subsection{\label{allen-crisp}Temporal Relationships}
%an explanation on allen's temporal relations.
%definition of fuzzy temporal interval
In this section, a brief introduction can be found, concerning \emph{temporal relationships}, sometimes also called `temporal relations'\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. Temporal relationships can be seen as relationships between temporal elements belonging to the same time domain. These relationships express how the temporal elements are related to one another, with respect to temporal precedence and overlap.
